A Large-Scale Study of Polymorphism in R

### 2019-04-05 paper/main.tex
more prose and some edits

\title{A Large-scale Study of Polymorphism in R}

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{?\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{37.6\%\xspace}

@@

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by \PROGRAMMERS programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

@@

{\bf L0} is limited in expressivity as it supports polymorphism for list in
a trivial way, it does not address some of the popular constructed data
types such as matrices and data frames, and cannot differentiate between
scalar values and vectors.  {\bf L1} types build on and extend the previous
level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. There is no subtyping relation between lists of different
types.

@@

{\bf L2} extends the previous system with more support for attributes.
While there is a finite number of results from the {\tt typeof} function,
this is not the case for \code{class} and \code{attributes}.  Classes and
attributes are user-defined.  Of course, not all classes are user-defined,
and R has a number of built-in or otherwise standard classes, ranging from
truly primitive classes such as {\tt numeric} to the more complex {\tt
  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
those classes and thus highlight user-defined ones.  As for attributes,
again some patterns arise naturally, for instance all matrices have a {\tt
  dims} attribute (indeed, it's this attribute that makes a matrix a
matrix).  Attributes are name-value pairs, to keep things simple we reduce
these to names. The only attribute that we treat specially is the {\tt
  class} for those we will recurse and capture the set of values defined by
the user (these are the class names for the object).  Thus, we extend {\bf
  L1} with attribute lists and classes: \attrclass{T}{C}{N} where {\tt T}
can be any L1 type other than data frame or matrix, {\tt C} and {\tt N}
stands for a list names.  The subtype relation is extended with
\attrclass{T}{C}{N}<:\attrclass{T}{C'}{D'} if {\tt C} $\subseteq$ {\tt C'}
and {\tt N}$\subseteq${\tt N'}.

@@

{\bf L2} works class and attribute information into its type signatures.
As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.

@@

\begin{table}[ht]
\label{tab:L2top10arg}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \sD & 46938 & 17.6\% \\ 
  \sL & 24113 & 9\% \\ 
  \sC & 21387 & 8\% \\ 
  \D & 20900 & 7.8\% \\ 
  \sF & 14556 & 5.5\% \\ 
  \sN & 14016 & 5.3\% \\ 
  unevaled & 9093 & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
  \C& 6939 & 2.6\% \\ 
  \attrclass{\l}{}{names} & 6137 & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\end{table}

@@

Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of polymorphism in {\bf L2}.
Indeed, our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
this would further reduce the amount of polymorphism.

\AT{Conclude L2}

@@

Given the data that we've collected, it's natural to wonder what the most polymorphic packages are.
Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
What is their purpose, and why are they so polymorphic?
Figure~\ref{tab:packagepolysize} has some answers.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\label{tab:packagepolysize}
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  knitLatex & 4 & 41 & \LaTeX \xspace generator from R data \\ 
  magrittr & 17 & 126 & Tidyverse pipe suite \\ 
  proto & 5 & 30 & Prototype object-based programming \\ 
  gnlm & 10 & 53 & ?? \\ 
  Formula & 17 & 82 & ?? \\ 
   \hline
\end{tabular}
\caption{5 most polymorphic packages.}
\end{table}

Now, what makes these so polymorphic?
\AT{maybe first score functions, penalizing big ones, and then consolidate?}

@@

	\iitem proto:

@@

@@

%
%
%
%
\subsection{Metaprogramming}

\AT{Make into subsection at the end of analysis? Or in global results, as a takeaway.}

Tidyverse owes some of its success to the new syntax defined by some of the included packages.
One such example is the pipe operator \%$>$\% that we saw in the last section, and indeed this function was highly exercised by many other packages \AT{get counts}.
We also grappled with the unfortunate truth that arguments to \%$>$\% can't always be evaluated, which causes some annoyances in the analysis.

We would like to know how often packages contain functions designed for metaprogramming.
As it happens, when arguments fail to evaluate during our analysis even as functions execute successfully, we are actually dealing with instances of metaprogramming.
Some functions, such as those in the magrittr package, will take one (or more) of their arguments and use them as an AST, performing surgery on the expression to create some new expression that they evaluate.
Sometimes, these arguments intended for metaprogramming {\it will} successfully evaluate in our tracer, but this is merely chance.
For example, consider the code snippet in Figure~\ref{fig:metaexworks}.

\begin{figure}[!hb]{\small\begin{lstlisting}[style=R]
mul_by_w_defaults <- function(x=1, y=1) x * y

4 %>% mul_by_w_defaults(2) == `%>%`(lhs = 4, rhs = mul_by(3))
\end{lstlisting}}\caption{Dates in R.}\label{fig:metaexworks}\end{figure}

In this example, our analysis will successfully evaluate the {\tt rhs} argument

@@

>>>>

\title{A Large-Scale Study of Polymorphism in R}

\newcommand{\PACKAGES}{11,463\xspace}
\newcommand{\PROGRAMMERS}{?\xspace}
\newcommand{\PERCENTCRAN}{83\%\xspace}
\newcommand{\CRANTOTAL}{13,841\xspace}
\newcommand{\RLOC}{15,050,267\xspace}
\newcommand{\CLOC}{9,373,542\xspace}
\newcommand{\YEARS}{20\xspace}
\newcommand{\INDEXCOINCIDENCE}{3,499\xspace}
\newcommand{\TOTALINDEXY}{6,561\xspace}
\newcommand{\INDEXYPERC}{53\%\xspace}		% TODO Not an accurate count.
\newcommand{\DATAPKGS}{206\xspace}
\newcommand{\DATAPKGSPERC}{1.5\%\xspace}
\newcommand{\METAARGCOUNT}{7,051\xspace}
\newcommand{\NUMPOLYTYPEMONOCLASS}{1,357\xspace}
\newcommand{\PERCSCALARMONO}{37.6\%\xspace}

@@

\begin{abstract}
The R programming language is widely used in a variety of scientific domains
for tasks related to data science. The language was designed to favor an
interactive style of programming with minimal syntactic and conceptual
overhead. This design is well suited to support interactive data analysis,
but is not well suited to generating performant code or catching programming
errors.  In particular, R has no type annotations and all operations are
dynamically checked at runtime. The starting point for our work is the
question: \emph{what could a static type system for R look like?}  To answer
that question we study the polymorphism that is present in \RLOC lines of R
code spread among some \PACKAGES packages, written over a
period of over \YEARS years by thousands of programmers.  We perform a dynamic
analysis, leveraging tests and use-cases, to determine the level of
polymorphism that is present in the code. We do this for several potential
notions of types. Our result suggest that polymorphism is important in some
key parts of the system but that relatively simple type annotations could be
used to capture most of the interesting cases.

@@

{\bf L0} is limited in expressivity as it supports polymorphism for list in
a trivial way, it does not address some of the popular constructed data
types such as matrices and data frames, and cannot differentiate between
scalar values and vectors.  {\bf L1} types build on and extend the previous
level with with scalar values for primitive types (\sI, \sD, \sC, \sL, \sR,
and \sX), type parametric lists (written \lT{T} for any type {\tt T}), data
frames (written \attr{\l}{df} as they are built up from lists of type any),
and matrices (\attr{\vec{\tt T}}{mat} where {\tt T} is a primitive type).
Subtyping is extended such that {\tt T}<:{$\vec{\tt T}$} for any primitive
type {\tt T}. The subtype relation is undefined for lists of different
types.

@@

{\bf L2} extends the previous system with more support for attributes.
While there is a finite number of results from the {\tt typeof} function,
this is not the case for \code{class} and \code{attributes}.  Classes and
attributes are user-defined.  Of course, not all classes are user-defined,
and R has a number of built-in or otherwise standard classes, ranging from
truly primitive classes such as {\tt numeric} to the more complex {\tt
  data.frame}.  We have already used the latter in {\bf L1}. We will ignore
those classes and thus highlight user-defined ones.  As for attributes,
again some patterns arise naturally, for instance all matrices have a {\tt
  dims} attribute (indeed, it's this attribute that makes a matrix a
matrix).  Attributes are name-value pairs, to keep things simple we reduce
these to names. The only attribute that we treat specially is the {\tt
  class} for those we will recurse and capture the set of values defined by
the user (these are the class names for the object).  Thus, we extend {\bf
  L1} with attribute lists and classes: \attrclass{T}{C}{A} where {\tt T}
can be any L1 type other than data frame or matrix, and {\tt C} and {\tt A}
are lists of classes and attribute names respectively.  The subtype relation is defined as:
\attrclass{T}{C}{A} $<:_{\mathbf{L2}}$ \xspace \attrclass{\dot T}{C'}{A'} if 
$ T <:_{\mathbf{L1}} \dot T$, $C \subseteq C'$, and $A \subseteq A'$.

@@

{\bf L2} works class and attribute information into its type signatures.
As per the discussion in Section~\ref{sec:L2def}, we omit builtin classes in order to highlight those which are user-defined.
{\bf L2} differs from {\bf L1} in the following ways:

	\item where {\bf L1} and {\bf L0} only considered classes and attributes when ascribing matrix or data frame types, {\bf L2} considers classes and attributes to be an integral part of the type system.
	Annotations in {\bf L2} include classes and attribute names when they are present.

	\item recall that subtyping between two {\bf L2} types is defined as

@@

\begin{table}[ht]
\label{tab:L2top10arg}
\centering
\begin{tabular}{lrl}
  \hline
Signature & Count & \% of All Signatures \\ 
  \hline
  \sD & 46938 & 17.6\% \\ 
  \sL & 24113 & 9\% \\ 
  \sC & 21387 & 8\% \\ 
  \D & 20900 & 7.8\% \\ 
  \sF & 14556 & 5.5\% \\ 
  \sN & 14016 & 5.3\% \\ 
  \ANY & 9093 & 3.4\% \\ 
  \attrclass{\D}{}{dim} & 6974 & 2.5\% \\ 
  \C& 6939 & 2.6\% \\ 
  \attrclass{\l}{}{names} & 6137 & 2.2\% \\ 
   \hline
\end{tabular}
\caption{Most frequent argument signatures in {\bf L2}.}
\end{table}

@@

Surprisingly, even with the added precision of classes and attributes, there is still a relatively small amount of argument-level polymorphism in {\bf L2}.
Our notion of subtyping on classes is rather weak, and one extension would be to include class hierarchies in the subtyping of classes;
this would further reduce the amount of polymorphism in this type system.

@@

Given the data that we've collected, it's natural to wonder what the most polymorphic packages are.
Indeed, we have shed some light on some of the most polymorphic {\it functions}, but what about the packages that these functions belong to?
What is their purpose, and why are they so polymorphic?
Figure~\ref{tab:packagepolysize} has some answers.
Note that we are in the context of {\bf L0}, here, to highlight current language patterns.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Thu Mar 28 18:15:26 2019
\begin{table}[ht]
\label{tab:packagepolysize}
\centering
\begin{tabular}{lrrr}
  \hline
  Package & \# of Functions &\# of Signatures & Description \\ 
  \hline
  knitLatex & 4 & 41 & \LaTeX \xspace generator from R data \\ 
  magrittr & 17 & 126 & Tidyverse pipe suite \\ 
  proto & 5 & 30 & Prototype object-based programming \\ 
  gnlm & 10 & 53 & Regression fit package \\ 
  Formula & 17 & 82 & ?? \\ 
   \hline
\end{tabular}
\caption{5 most polymorphic packages (in {\bf L0}).}
\end{table}

Now, what makes these so polymorphic?

\AT{TODO if time: fix to score per fun, then tally per package.}

@@

	\item proto: most of the polymorphism in this package comes from its implementation of assignment ({\tt \$<-}) for proto class objects.
	It's a small, 3 argument function which is exceedingly polymorphic, with over 20 unique call signatures.
	As theoretically any R object can have a proto class, it stands to reason that this function ought to be quite polymorphic indeed.

	\item

	\item

@@

It's highly improbable that a static type system designed for an untyped language will be able to capture all of the language's existing code (without resorting to an \ANY type or abusing unions).
It's also unclear how useful a type system would be if it {\it did} capture everything.
Indeed, many of the packages discussed here are really designed to take in anything for particular arguments (e.g., left-hand side of \%$>$\%), and restricting these functions should not be the goal of any static type system designed after a language has matured.

@@

%
%
%
%
\subsubsection{Metaprogramming and Unevaluated Arguments}

Our analysis uncovered a non-negligible amount of argument positions which were {\it never} evaluated across all observed function calls.
This is indicative of a desire to metaprogram arguments and otherwise deconstruct their AST, as the \%$>$\% function does.
This is an incredibly powerful language feature, and may well be at the source of R's popularity, if in a roundabout way.
Tidyverse, by far the most popular package suite in R, owes some of its success to the new syntax defined by some of the included packages, and it's this ability to metaprogram arguments which allowed Tidyverse to define its unique syntax.
An example of that syntax is \%$>$\%, and indeed this function was highly exercised by over 100 other packages.

%
%
%
%
%
%
\section{Discussion}

\AT{Short section to recap the analysis results, discuss trends, consolidate our suggestions, talk about argument vs. function poly.}

@@

\AT{Retool this into possible meta discussion after results}
